---
title: "ASMFC MSE Workshop: Alternative Operating Models"
author: "Gavin Fay"
date: "August 2021"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Operating Model uncertainty

In the previous example we characterized OM uncertainty using paramter uncertainty in the fit to the original data.

Here we will compare the performance of our simple HCRs given operating model structure uncertainty.

1. Build a function for when the operating model dynamics are governed by the Pella-Tomlinson model (i.e. yield function is not symmetric)
2. Compare the performance of our HCRs from day 1 to both OMs


_Stretch-goal_ 

3. Develop an OM that includes process error (recruitment variabilty?) in the population dynamics


#### first load packages
 
```{r, results='hide'}
#install.packages('ggplot2')
library(ggplot2)
library(Hmisc)
library(mvtnorm)
```

#### load the functions we used yesterday
(sourced from a script to make things easy and reduce clutter in this document)
```{r}
source("first-mse-functions.R")
```

#### The Operating Model  

The population dynamics for the operating model (the 'real' dynamics) are governed by the Pella-Tomlinson equation:  
$$ B_{y+1} = B_y + B_y * r * \left(1 - \left(\frac{B_y}{K} \right)^{m-1} \right) - C_y $$
where $B_y$ is the biomass in year $y$,
$C_y$ is the catch in year $y$,
$r$ is the population intrinsic growth rate,
$K$ is the population carrying capacity, and
$m$ is a shape parameter that determines the location of BMSY.  

First task is to create a function for the P-T dynamics
(hint modify the schaefer function that we used yesterday. Remember there is an extra parameter)
```{r}
pellat <- function(B,C,K,r,m) {
  #function schaefer takes the current biomass, a catch, 
  #and the model parameters to compute next year's biomass
  res <- B + B * r * (1 - (B/K)^(m-1)) - C
  return(max(0.001,res))  # we add a constraint to prevent negative biomass
}
```

Once you have this, condition your OM as yesterday but with a fixed value for the shape parameter of 1.2
(Hint, you will want to adjust the biomass projection to use your new function)

change the dynamics function

```{r}
#biomas projection
dynamics <- function(pars,C,yrs,m) {
  # dynamics takes the model parameters, the time series of catch, 
  # & the yrs to do the projection over
  
  # first extract the parameters from the pars vector (we estimate K in log-space)
  K <- exp(pars[1])
  r <- exp(pars[2])
  
  # find the total number of years
  nyr <- length(C) + 1
  
  # if the vector of years was not supplied we create 
  # a default to stop the program crashing
  if (missing(yrs)) yrs <- 1:nyr
  
  #set up the biomass vector
  B <- numeric(nyr)
  
  #intialize biomass at carrying capacity
  B[1] <- K
  # project the model forward using the schaefer model
  for (y in 2:nyr) {
    B[y] <- pellat(B[y-1],C[y-1],K,r,m)
  }
  
  #return the time series of biomass
  return(B[yrs])
  
  #end function dynamics
}  

```

#####Specify input data and associated years
```{r}
data.years <- 1991:2013
harvest <- c(0.1,3,15,52,76,139,95,93,84,93,86,103,104,
             92,46,67,59,30,54,59,47,33,44)
index <- c(NA,NA,NA,NA,NA,NA,NA,NA,935,NA,1057,NA,678,NA,
           420,NA,554,NA,458,NA,474,NA,280)
```

```{r}
# function to calculate the negative log-likelihood
nll <- function(pars,C,U,m) {  #this function takes the parameters, the catches, and the index data
  sigma <- exp(pars[3])  # additional parameter, the standard deviation of the observation error
  B <- dynamics(pars,C,m=m)  #run the biomass dynamics for this set of parameters
  Uhat <- B   #calculate the predicted biomass index - here we assume an unbiased absolute biomass estimate
  output <- -sum(dnorm(log(U),log(Uhat),sigma,log=TRUE),na.rm=TRUE)   #calculate the negative log-likelihood
  return(output)
#end function nll
}
```


Function to perform the assessment and estimate the operating model parameters  
(i.e. to fit the logistic model to abundance data)
```{r}
assess <- function(catch,index,calc.vcov=FALSE,pars.init,m=2) {
  # assess takes catch and index data, initial values for the parameters,
  # and a flag saying whether to compute uncertainty estimates for the model parameters
  
  #fit model
  # optim runs the function nll() repeatedly with differnt values for the parameters,
  # to find the values that give the best fit to the index data
  res <- optim(pars.init,nll,C=catch,U=index,m=m,hessian=TRUE)
  
  # store the output from the model fit
  output <- list()
  output$pars <- res$par
  output$biomass <- dynamics(res$par,catch,m=m)
  output$convergence <- res$convergence
  output$nll <- res$value
  if (calc.vcov)
    output$vcov <- solve(res$hessian)
  
  return(output)
#end function assess
}
```

Now we have written the functions to do the calculations, we can run them and perform the assessment. 

First define initial parameter vector for: log(K), r, sigma  
```{r}
ini.parms <- c(log(1200), log(0.1), log(0.3))
```

Fit the logistic model to data:  
```{r}
redfish <- assess(harvest,index,calc.vcov=TRUE,ini.parms,m=1.2)
redfish
```

Extract the maximum likelihood and parameter estimates  
```{r}
biomass.mle <- redfish$biomass
print(biomass.mle)
pars.mle <- redfish$pars
print(pars.mle)
```


To obtain plausible alternatives for the parameters of the operating model, we will use the statistical uncertainty from the estimation by sampling parameter sets from the estimated variance-covariance matrix.  
```{r}
m=1.2
#define the number of iterations for the MSE 
niter <- 500 
#set up a storage matrix for our alternative parameter sets
pars.iter <- matrix(NA,nrow = niter, ncol=3) 
colnames(pars.iter) <- c("logK","r","sigma")

# generate the sets of parameter values
for (i in 1:niter) {
  pars.iter[i,] <- mvtnorm::rmvnorm(1, mean = redfish$pars,
                       sigma = redfish$vcov)
}

# Now generate replicate model outputs
biomass.iter <- data.frame()
for (i in 1:niter) {
  #here we calculate the biomass trajectory for each of the above sampled parameter vectors
  biomass.iter <- rbind(biomass.iter,
                        data.frame(year = seq(min(data.years),
                                              max(data.years)+1),
                                   biomass = dynamics(pars.iter[i,], harvest,m=m),
                                   iter = i))
}
```


We can now plot the estimated biomass time series
```{r}
Fig1 <- ggplot(data=biomass.iter,aes(x=year,y=biomass))
Fig1 + 
     stat_summary(fun.data = "median_hilow",
                    fun.args = list(conf.int=1),
                    geom = "ribbon" ,
                    alpha=0.5,
                    fun.min = function(x)0,
                    fill = "gray")  + 
     stat_summary(fun.data = "median_hilow",
                    geom = "smooth" ,
                    color = gray(0.6))  +   
     geom_line(aes(y=harvest,x=year), data = data.frame(harvest = harvest, 
               year = data.years),lty=2) +
     geom_point(aes(y=index, x=year), data = data.frame(index=index, 
               year = data.years)) +
     geom_line(aes(y=biomass,x=year,group=iter,col=iter),data = subset(biomass.iter,iter%in%1:10)) +
      ylab("Estimated B and C (million tonnes)") + 
  theme_bw() +
  guides(col=FALSE)
  
```

The shaded area indicates the range of the biomass time series, with the dark line the median. The lighter lines indicate indiviudal biomass trajectories.  



#### Applying the Management Strategy  

We have now conditioned our operating model. We will conduct the MSE loop over a 20 year projection period, with the catches set each year by repeated estimation of the current biomass and application of a harvest control rule.  

Define the years for the projection:  
```{r}
proj.years <- 2014:2034
```


##### Data generation

We write a function to generate the observations (new biomass index values) from the operating model.  
```{r}
observe <- function(biomass, sigma) {
  biomass * exp(rnorm(1, -0.5*sigma^2, sigma))
}
```
This function takes the true biomass from the operating model, and generates the data by adding (lognormally distributed) observation error.    


##### Harvest Control Rule

We first demonstrate the MSE using a fixed target exploitation rate - the control rule calculates the catch for next year based on a fixed percentage (10%) of the most recent biomass estimate.  
```{r}
control.pars <- list()
control.pars$Htarg <- 0.1
control <- function(estimated.biomass, control.pars) {
  control.pars$Htarg
}
```

We assume perfect implementation of the strategy - in that the realized catch is the same as the TAC.  
```{r}
implement <- function(TAC,...) {
  TAC
}
```


 
```{r, include=FALSE}
evaluate <- function(pars.iter, biomass.iter,
                     control.pars, data.years, proj.years,
                     iterations,m=2,...) {
    # function arguments:
    # pars.iter & biomass.iter, the parameters & historical biomass trajectories of the operating model
    # control.pars, the specifications of the harvest control rule
  
    # set up some indexing values
    iyr <- length(data.years)+1
    pyr <- length(proj.years)
    yrs <- c(data.years, proj.years, max(proj.years)+1)
    
    # set up a data frame to store the results
    res <- data.frame()
    
    # loop over the iterations of the MSE, each iteration conducts a 20 year projection with annual generation of biomass    
    # observations and appliations of the control rule.
    for(i in 1:iterations) {
      
      #extract the parameters for this iteration
      K.i <- exp(pars.iter[i,1])
      r.i <- exp(pars.iter[i,2])
      sig.i <- exp(pars.iter[i,3])
      
      #set up vectors for time series of interest.
      biomass.i <- c(subset(biomass.iter, iter==i)$biomass, numeric(pyr))
      index.i <- c(index,numeric(pyr))
      catch.i <- c(harvest, numeric(pyr))
      TAC.i <- numeric(pyr)
      
      # loop over the projection period.
      for (y in iyr:(iyr+pyr-1)) {
        #generate the data for the most recent year
        index.i[y] <- observe(biomass.i[y] , sig.i)
        #calculate the TAC based on the harvest control rule
        # note that the control rule ONLY sees the index data, not the operating model biomass.
        TAC.i [y]  <- control(index.i[y], control.pars) * index.i[y]
        #find the realized catch after implementation error
        catch.i[y] <- implement(TAC.i[y],...)
        
        # update the true biomass of the operating model based on the output of the HCR
        biomass.i[y+1] <- pellat(biomass.i[y],catch.i[y],K.i,r.i,m=m)
      
      #end projection year loop for iteration i  
      }
     
      #store the results for this iteration
      res <- rbind(res, data.frame(year = yrs[-length(yrs)],
                                   value = index.i, type = "index", iter = i),
                   data.frame(year = yrs[-length(yrs)],
                              value = catch.i, type = "catch", iter=i),
                   data.frame(year = yrs, value = biomass.i,
                              type= "biomass", iter=i),
                   data.frame(year = yrs[-length(yrs)],
                              value = TAC.i, type = "tac", iter=i)) 
    #end loop over iterations
    }
    return(res)
#end function evaluate()
}
```


Project with fixed 10% exploitation rate of estimated biomass for all iterations & 20 yrs  
```{r}
project.fixed <- evaluate(pars.iter, biomass.iter, control.pars, data.years,
                          proj.years, niter)
tail(project.fixed)
```



Plot the projection:  
```{r}
projection.plot(dplyr::filter(project.fixed, type != "tac"))
```



#### Management using alternative harvest control rules

Define a HCR that converts estimated biomass into a harvest rate using a functional form determined by values in 'control.pars'.  
```{r}
control <- function(estimated.biomass, control.pars) {
  H1 <- control.pars$H1
  H2 <- control.pars$H2
  Bmax <- control.pars$Bmax
  B2 <- control.pars$B2
  B1 <- control.pars$B1
  
  harv <- ifelse(estimated.biomass >= B1, H1,
                 ifelse(estimated.biomass < B2, H2,
                        (H1-H2)/(B1-B2)*(estimated.biomass - B2) + H2))
  
  return(harv)
  
#end function control
}
```


Define control parameters for HCR using reference points  
We (arbitrarily) set the threshold and limit biomass reference points as 50% & 20% of the maximum observed survey index value during the historical period.  
The target exploitation rate is set at 5%.  
```{r}
control.pars <- list()
control.pars$H1 <- 0.05
control.pars$H2 <- 0
control.pars$Bmax <- max(index, na.rm =TRUE)
control.pars$B2 <- 0.2*control.pars$Bmax
control.pars$B1 <- 0.5*control.pars$Bmax
```


Plot the HCR shape:  
```{r}
plot(c(0,control.pars$B2,control.pars$B1,control.pars$Bmax),
     c(control.pars$H2,control.pars$H2,control.pars$H1,control.pars$H1),
     type='l',axes=F,xlab="estimated biomass",ylab="exploitation rate",
     ylim=c(0,1.2*control.pars$H1))
axis(1,at=c(control.pars$B2,control.pars$B1),labels=c("B2","B1"))
axis(2,at=c(control.pars$H2,control.pars$H1),labels=c("H2","H1"))
box()
```


Conduct the evaluation by projecting system forward in time
```{r}
project.hcr.sf <- evaluate(pars.iter, biomass.iter, control.pars,
                        data.years, proj.years, niter, m = 2)
```
Plot the trajectories:  
```{r}
projection.plot(project.hcr.sf)
```

Conduct the evaluation by projecting system forward in time
```{r}
project.hcr.pt <- evaluate(pars.iter, biomass.iter, control.pars,
                        data.years, proj.years, niter, m = 1.2)
```
Plot the trajectories:  
```{r}
projection.plot(project.hcr.pt)
```


#### Diagnostics

We have run the evaluations for 4 HCRs. We can now compare these.  
Create an object containing all the results:    
```{r}
MSE <- rbind(cbind(project.hcr.sf, HCR="schaefer",
                   implement = "no overshoot"),
             cbind(project.hcr.pt, HCR="pella-t",
                   implement = "no overshoot"))
head(MSE)
```

Summarize biomass & catch for all 4 options:  
```{r}
Fig5 <- ggplot(data=subset(MSE, type !="index" &
                          year %in% proj.years), 
               aes(x=HCR, y=value, ymin=0))
Fig5 + geom_boxplot(aes(fill=implement), width = 1) + facet_wrap(~type, scale="free_y") + ylab("Million tonnes") + scale_fill_grey(start=0.5) + theme_bw()
```


#### Performance statistics
```{r}
#Yield based metrics  (e.g. average annual catch)
#Stock Biomass metrics (e.g. distribution for B/BMSY, P(B>BLIM), etc.)
#Inter-annual stability of catch advice (e.g. how often the control rule closes the fishery)

aac2 <- with(MSE[MSE$year>max(data.years) & MSE$type=="catch",],
            aggregate(value,by=list(iter=iter,HCR=HCR,implement=implement),FUN=mean,na.rm=TRUE))


Fig6 <- ggplot(data=subset(aac2), 
               aes(x=HCR, y=x, ymin=0))
Fig6 + geom_boxplot(aes(fill=implement), width = 1)  + ylab("Million tonnes") + scale_fill_grey(start=0.5) + theme_bw()

# years B > BLIM
# BLIM = 0.25*K  (we specify BLIM for our performance as half BMSY)
blim <- 0.25*exp(pars.iter[,1])

num.above <- function(vec,threshold) {
 length(vec[vec>threshold])/length(vec)
}
MSE$blim <- blim[MSE$iter]
MSE$above.blim <- ifelse(MSE$value>MSE$blim,1,0)

above.blim <- with(MSE[MSE$year>max(data.years) & MSE$type=="biomass",],
            aggregate(above.blim,by=list(iter=iter,HCR=HCR,implement=implement),FUN=sum,na.rm=TRUE))
above.blim$x <- above.blim$x/length(proj.years)

Fig7 <- ggplot(data=subset(above.blim), 
               aes(x=HCR, y=x, ymin=0))
Fig7 + geom_boxplot(aes(fill=implement), width = 1)  + ylab("Proportion of years above BLIM") + scale_fill_grey(start=0.5) + theme_bw()

# num years fishery is open
not.closed <- with(MSE[MSE$year>max(data.years) & MSE$type=="catch",],
             aggregate(value,by=list(iter=iter,HCR=HCR,implement=implement),FUN=num.above,threshold=0))

Fig8 <- ggplot(data=subset(not.closed), 
               aes(x=HCR, y=x, ymin=0))
Fig8 + geom_boxplot(aes(fill=implement), width = 1)  + ylab("Proportion of years TAC > 0") + scale_fill_grey(start=0.5) + theme_bw()

```


