---
title: "ASMFC MSE Workshop: Model-based Estimation Model"
author: "Gavin Fay"
date: "August 2021"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Replacing our assessment with an estimation model

In the previous examples we used our survey index as the estimate of abundance.

Here we will instead use a model-based estimator, and fit a Schaefer model to the data from the OM to estimate MSY reference points for use in the control rule.
(hint we can reuse the code we developed to condition the OM)

1. Build a function for fitting an estimation model to the index.i time series (don't fit to the future time steps that haven't happened yet)
2. Calculate the TAC based on the last year's (model-estiamted) biomass and the estimate of FMSY (r/2) from the fitted model.
3. run a FMSY based control rule

_Stretch-goal_ 

4. Compare the performance of the EM-based strategy when the OM is the Pella-Tomlinson model.


#### first load packages
 
```{r, results='hide'}
#install.packages('ggplot2')
library(ggplot2)
library(Hmisc)
library(mvtnorm)
```

#### load the functions we used yesterday
(sourced from a script to make things easy and reduce clutter in this document)
```{r}
source("first-mse-functions.R")
```

First task is to create a function for the estimation moodel that will use the index data and call the 'assess()' function.
```{r, include=FALSE}
em_evaluate <- function(pars.iter, biomass.iter,
                     control.pars, data.years, proj.years,
                     iterations, ...) {
    # function arguments:
    # pars.iter & biomass.iter, the parameters & historical biomass trajectories of the operating model
    # control.pars, the specifications of the harvest control rule
  
    # set up some indexing values
    iyr <- length(data.years)+1
    pyr <- length(proj.years)
    yrs <- c(data.years, proj.years, max(proj.years)+1)
    
    # set up a data frame to store the results
    res <- data.frame()
    
    # loop over the iterations of the MSE, each iteration conducts a 20 year projection with annual generation of biomass    
    # observations and appliations of the control rule.
    for(i in 1:iterations) {
      
      #extract the parameters for this iteration
      K.i <- exp(pars.iter[i,1])
      r.i <- pars.iter[i,2]
      sig.i <- pars.iter[i,3]
      
      #set up vectors for time series of interest.
      biomass.i <- c(subset(biomass.iter, iter==i)$biomass, numeric(pyr))
      index.i <- c(index,numeric(pyr))
      catch.i <- c(harvest, numeric(pyr))
      TAC.i <- numeric(pyr)
      
      # loop over the projection period.
      for (y in iyr:(iyr+pyr-1)) {
        #generate the data for the most recent year
        index.i[y] <- observe(biomass.i[y] , sig.i)
        #calculate the TAC based on the harvest control rule
        # note that the control rule ONLY sees the index data, not the operating model biomass.
        
        # Run a Schaefer production model as the Estimation model
        assess.i <- assess(catch.i[1:y-1],index.i[1:y-1],calc.vcov=TRUE,ini.parms)
        
        # Make the TAC a function of the estimation model results & FMSY (r/2)
        TAC.i[y] <- assess.i$biomass[y]*(exp(assess.i$pars[2])/2)
        
        #TAC.i [y]  <- control(index.i[y], control.pars) * index.i[y]
        #find the realized catch after implementation error
        catch.i[y] <- implement(TAC.i[y])
        
        # update the true biomass of the operating model based on the output of the HCR
        biomass.i[y+1] <- schaefer(biomass.i[y],catch.i[y],K.i,r.i)
      
      #end projection year loop for iteration i  
      }
     
      #store the results for this iteration
      res <- rbind(res, data.frame(year = yrs[-length(yrs)],
                                   value = index.i, type = "index", iter = i),
                   data.frame(year = yrs[-length(yrs)],
                              value = catch.i, type = "catch", iter=i),
                   data.frame(year = yrs, value = biomass.i,
                              type= "biomass", iter=i),
                   data.frame(year = yrs[-length(yrs)],
                              value = TAC.i, type = "tac", iter=i)) 
    #end loop over iterations
    }
    return(res)
#end function evaluate()
}
```

load data
```{r}
data.years <- 1991:2013
harvest <- c(0.1,3,15,52,76,139,95,93,84,93,86,103,104,
             92,46,67,59,30,54,59,47,33,44)
index <- c(NA,NA,NA,NA,NA,NA,NA,NA,935,NA,1057,NA,678,NA,
           420,NA,554,NA,458,NA,474,NA,280)
```

Condition the OM
```{r}
set.seed(8675309)
ini.parms <- c(log(1200), log(0.1), log(0.3))
redfish <- assess(harvest,index,calc.vcov=TRUE,ini.parms)
biomass.mle <- redfish$biomass
print(biomass.mle)
pars.mle <- redfish$pars
#define the number of iterations for the MSE 
niter <- 500 
#set up a storage matrix for our alternative parameter sets
pars.iter <- matrix(NA,nrow = niter, ncol=3) 
colnames(pars.iter) <- c("logK","r","sigma")

# generate the sets of parameter values
for (i in 1:niter) {
  pars.iter[i,] <- mvtnorm::rmvnorm(1, mean = redfish$pars,
                       sigma = redfish$vcov)
}

# Now generate replicate model outputs
biomass.iter <- data.frame()
for (i in 1:niter) {
  #here we calculate the biomass trajectory for each of the above sampled parameter vectors
  biomass.iter <- rbind(biomass.iter,
                        data.frame(year = seq(min(data.years),
                                              max(data.years)+1),
                                   biomass = dynamics(pars.iter[i,], harvest),
                                   iter = i))
}
```

#### Applying the Management Strategy  

Define the years for the projection:  
```{r}
proj.years <- 2014:2034
```

conduct the evaluation with the model-based estimation method
```{r}
project.emhcr <- em_evaluate(pars.iter, biomass.iter, control.pars,
                        data.years, proj.years, 100)
```

Plot the trajectories:  
```{r}
projection.plot(project.emhcr)
```


